### 2020-01-27 manuscript.tex
Protocol order rearrangement

@@

% REMEMBER to define NNI and SPR acronyms in introduction preferably
\subsection*{Narrow Exchange Rate} \label{sect:NER}

@@

Lakner et al. 2008 \cite{lakner2008efficiency} found that tree operators which perterb topology (such as NNI and SPR) consistently perform better than those which also change branch lengths (such as LOCAL \cite{simon1998local} and Continuous Change \cite{jow2002bayesian}). If Narrow Exchange was adapted to the relaxed clock model by ensuring that genetic distances remain constant after the proposal, its performance may be improved even further.

@@

Here, we present the \textbf{Narrow Exchange Rate} (NER) operator. Let $r_A$, $r_B$, $r_C$, and $r_D$ be the clock rates of nodes $A$, $B$, $C$, and $D$, respectively. In addition to the modest topological change applied by Narrow Exchange, NER also proposes new clock rates ${r_A}^\prime$, ${r_B}^\prime$, ${r_C}^\prime$, and ${r_D}^\prime$. NER is compatible with both the \textit{real} and \textit{quant} parameterisations. Analagous to the Constant Distance operator, these rates are proposed such that genetic distances between nodes $A$, $B$, $C$, and $E$ are maintained. Thus, there are $\binom{4}{2} = 6$ pairwise distance constraints.

@@

The 48 operators generated by this pipeline are evaluated and compared in \textbf{\nameref{sect:results}}.

@@

\begin{align}
	\alpha \sim \text{Bactrian}(m) \equiv \frac{1}{2}\text{Normal}(-m, 1-m^2) + \frac{1}{2}\text{Normal}(m, 1-m^2)
\end{align}

where $0 \leq m < 1$ describes the modality of the Bactrian distribution. When $m=0$, the Bactrian distribution is equivalent to a $\text{Normal}(0, 1)$ distribution. As $m \rightarrow 1$, the distribution becomes increasingly bimodal (\textbf{Fig. XXX}). Yang et al. 2013 \cite{yang2013searching} suggest that $\text{Bactrian}(m=0.95)$ yields a proposal kernel superior to the uniform kernel by placing minimal probability on step which are too small or too large.

 %The bimodally shaped Bactrian distribution ensures that most step sizes are of an intermediate magnitude.

%In this article we compare the performance the uniform and Bactrian proposal kernels in the clock model.

\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l|} 
 \hline
  &&& \\
 & Operator & Proposal & Parameters  \\
 &&& \\
 1 & \textbf{Random walk operator} & $x^\prime \leftarrow x + w\alpha$ & $t, \sigma, r, q$ \\
 2 & \textbf{Scale operator} & $x^\prime \leftarrow $ & $\sigma, r$  \\
 %3 & \textbf{Uniform operator} & \begin{align} a \\ b \end{align} & $t, q$  \\
 4 & \textbf{Constant distance operators} & $x^\prime \leftarrow x + w\alpha$ & $t$ \\
 5 & \textbf{NER operator} & $x^\prime \leftarrow x + w\alpha$ & $t$ \\
 \hline
\end{tabular}
\caption{Proposal kernels}
\label{table:kernels}
\end{table}

@@

To avoid a cross-product explosion, the three targets for clock model improvement (\textbf{\nameref{sect:rateparams}}, \textbf{\nameref{sect:NER}}, and \textbf{\nameref{sect:randomwalks}}) are evaluated sequentially, in the order presented in this paper. The setting(s) which are considered to be the best in each step are then incorporated into the following step. This protocol and its outcomes are summarised in \textbf{Fig. \ref{fig:tournament}}.

@@

17  &  78  &  1/4/8/16  &  1/2.6/8/14.1  &  0.6/1.7/6.4/10.4  &  Lartillot 2012 \cite{Lartillot_2012} \\

18  &  79  &  1/4/8/16  &  0.1/1.3/0.8/4.5  &  0/0.1/0.1/0.8  &  Oaks 2011 \cite{Oaks_2011} \\

19  &  94  &  1/4/8/11  &  0.1/1.8/3/3.7  &  0.1/0.8/1.3/1.7  &  Rightmyer 2013 \cite{Rightmyer_2013} \\

20  &  106  &  1/4/8/16  &  0.6/3.1/5.1/11.6  &  0.3/1.2/2/4.5  &  Moyle 2016 \cite{Moyle_2016} \\

21  &  110  &  1/4/8/16  &  0.5/2.1/3.8/7.1  &  0.3/1.4/2.5/4.3  &  Fong 2012 \cite{Fong_2012} \\

22  &  145  &  1/2  &  1.6/3  &  0.5/1  &  Anderson 2013 \cite{Anderson_2013} \\

23  &  152  &  1/4/5  &  1/3.5/3.6  &  0.6/1.6/1.6  &  Day 2013 \cite{Day_2013} \\

24  &  187  &  1/4/8/16  &  0.3/0.9/1.4/3.5  &  0.2/0.7/1.1/2.9  &  Branstetter 2017 \cite{Branstetter_2017} \\

25  &  197  &  1/4/8/14  &  0.9/3.1/6.1/11.6  &  0.3/1.6/3.5/6.3  &  Horn 2014 \cite{Horn_2014} \\

26  &  235  &  1/4/8/16  &  2.3/6.4/8.5/25.7  &  0.9/2.3/3.2/12.2  &  Reddy 2017 \cite{Reddy_2017} \\

27  &  237  &  1/4/5  &  0.8/2.4/3.1  &  0.2/1.2/1.5  &  Murray 2013 \cite{Murray_2013} \\

28  &  608  &  1/4/8/10  &  0.9/3.7/7.1/8.6  &  0.7/2.8/5.4/6.4  &  Near 2013 \cite{Near_2013} \\

29  &  1171  &  1/4  &  0.8/3.9  &  0.8/2.4  &  Looney 2016 \cite{Looney_2016} \\

@@

>>>>

\renewcommand{\arraystretch}{1.3}

@@

\begin{align}
	\Sigma \sim \text{Bactrian}(m) \equiv \frac{1}{2}\text{Normal}(-m, 1-m^2) + \frac{1}{2}\text{Normal}(m, 1-m^2)
\end{align}

where $0 \leq m < 1$ describes the modality of the Bactrian distribution. When $m=0$, the Bactrian distribution is equivalent to a $\text{Normal}(0, 1)$ distribution. As $m \rightarrow 1$, the distribution becomes increasingly bimodal (\textbf{Fig. \ref{fig:bactrian}}). Yang et al. 2013 \cite{yang2013searching} suggest that $\text{Bactrian}(m=0.95)$ yields a proposal kernel superior to the uniform kernel, by placing minimal probability on steps which are too small or too large.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/bactrian.pdf}
\caption{\textbf{The Bactrian proposal kernel.} Y-axis corresponds to probability density $f(\Sigma|m)$.  }
\label{fig:bactrian}
\end{figure}

In this article we compare the performance the uniform and Bactrian proposal kernels in the clock model. Three Bactrian distributions are compared($m=0.92$, $m=0.95$, and $m=0.98$). The clock model operators which are augmented by the Bactrian distribution are described in \textbf{Table \ref{table:kernels}}.

\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l|} 
 \hline
 & Operator & Proposal & Parameters  \\
 1 & \textbf{Random walk operator} & $x^\prime \leftarrow x + s\Sigma$ & $t, \sigma, r, q$ \\
 2 & \textbf{Scale operator} & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\sigma, r$  \\
 %3 & \textbf{Uniform operator} & $\begin{array} {rl} y &\leftarrow s\Sigma \times \frac{u - x}{x - l} \\ x^\prime &\leftarrow \frac{u + l * y}{y + 1}  \end{array}$ & $t, q$  \\
 3 & \textbf{Constant distance operators} & $x^\prime \leftarrow x + s\Sigma$ & $t$ \\
 \hline
\end{tabular}
\caption{Summary of proposal kernels $q(x^\prime|x)$ of clock model operators. In each operator, $\Sigma$ is drawn from either a $\text{Noramlised-Bactrian}(m)$ or $\text{Uniform}(-\sqrt{3},\sqrt{3})$ distribution. The scale size $s$ is tunable. The proposal kernel may apply to node heights $t$, clock standard deviation $\sigma$, clock rates $r$ (\textit{real} only), and clock rate quantiles $q$ (\textit{quant} only). The Scale operator acts on parameters with non-negative domains. }
\label{table:kernels}
\end{table}

\newpage
% REMEMBER to define NNI and SPR acronyms in introduction preferably
\subsection*{Narrow Exchange Rate} \label{sect:NER}

@@

Lakner et al. 2008 \cite{lakner2008efficiency} found that tree operators which perterb topology (such as NNI and SPR) consistently perform better than those which also change branch lengths (such as LOCAL \cite{simon1998local} and Continuous Change \cite{jow2002bayesian}). If Narrow Exchange was adapted to the relaxed clock model by ensuring that genetic distances remain constant after the proposal, its performance may be improved even further, which may in turn permit proposing a new node height $t_D$ and therefore changing branch time lengths (without affecting branch distance lengths).

@@

Here, we present the \textbf{Narrow Exchange Rate} (NER) operator. Let $r_A$, $r_B$, $r_C$, and $r_D$ be the clock rates of nodes $A$, $B$, $C$, and $D$, respectively. In addition to the modest topological change applied by Narrow Exchange, NER also proposes new clock rates ${r_A}^\prime$, ${r_B}^\prime$, ${r_C}^\prime$, and ${r_D}^\prime$. We also consider a case of the NER operator which proposes a new node height:

\begin{align}
	{t_D}^\prime \leftarrow t_D + \Sigma
\end{align}

for random walk step size $\Sigma$ drawn from a uniform or \textbf{\nameref{sect:randomwalks}}. NER is compatible with both the \textit{real} and \textit{quant} parameterisations. Analagous to the Constant Distance operator, these rates are proposed such that genetic distances between nodes $A$, $B$, $C$, and $E$ are maintained. Thus, there are $\binom{4}{2} = 6$ pairwise distance constraints.

@@

The 48 operators generated by this pipeline are evaluated and compared in \textbf{\nameref{sect:results}}. Each operator is considered with and without a random walk on $t_D$ and thus there are 96 total settings.

@@

@@

To avoid a cross-product explosion, the three targets for clock model improvement (\textbf{\nameref{sect:rateparams}}, \textbf{\nameref{sect:randomwalks}}, and \textbf{\nameref{sect:NER}}) are evaluated sequentially, in the order presented in this paper. The setting(s) which are considered to be the best in each step are then incorporated into the following step. This protocol and its outcomes are summarised in \textbf{Fig. \ref{fig:tournament}}.

@@

17  &  79  &  1/4/8/16  &  0.1/1.3/0.8/4.5  &  0/0.1/0.1/0.8  &  Oaks 2011 \cite{Oaks_2011} \\

18  &  94  &  1/4/8/11  &  0.1/1.8/3/3.7  &  0.1/0.8/1.3/1.7  &  Rightmyer 2013 \cite{Rightmyer_2013} \\

19  &  106  &  1/4/8/16  &  0.6/3.1/5.1/11.6  &  0.3/1.2/2/4.5  &  Moyle 2016 \cite{Moyle_2016} \\

20  &  110  &  1/4/8/16  &  0.5/2.1/3.8/7.1  &  0.3/1.4/2.5/4.3  &  Fong 2012 \cite{Fong_2012} \\

21  &  152  &  1/4/5  &  1/3.5/3.6  &  0.6/1.6/1.6  &  Day 2013 \cite{Day_2013} \\

22  &  187  &  1/4/8/16  &  0.3/0.9/1.4/3.5  &  0.2/0.7/1.1/2.9  &  Branstetter 2017 \cite{Branstetter_2017} \\

23  &  197  &  1/4/8/14  &  0.9/3.1/6.1/11.6  &  0.3/1.6/3.5/6.3  &  Horn 2014 \cite{Horn_2014} \\

24  &  235  &  1/4/8/16  &  2.3/6.4/8.5/25.7  &  0.9/2.3/3.2/12.2  &  Reddy 2017 \cite{Reddy_2017} \\

25  &  237  &  1/4/5  &  0.8/2.4/3.1  &  0.2/1.2/1.5  &  Murray 2013 \cite{Murray_2013} \\

26  &  608  &  1/4/8/10  &  0.9/3.7/7.1/8.6  &  0.7/2.8/5.4/6.4  &  Near 2013 \cite{Near_2013} \\

@@

We compared the three rate parameterisations described in \textbf{\nameref{sect:rateparams}}. All three settings use the standard BEAST2 clock model operators from Drummond et al. 2006 \cite{drummond2006relaxed}. \textit{real} and \textit{quant} additionally use the constant-distance tree operators described by Zhang and Drummond 2020. To determine whether the difference in performance between \textit{real}/\textit{quant} versus \textit{cat} is because of the constant-distance tree operators or the parameterisation itself, we also included benchmarked two additional settings: \textit{real 2006} and \textit{quant 2006}, which do not used the constant-distance operators. These five settings are validated in \textbf{\nameref{sect:WCSS_appendix}}.

\textbf{Fig. \ref{fig:parameterisationResults}} shows that the \textit{real 2006} performs considerably worse than any of the other settings. This is due to the poor sampling of the prior under this setting (ie. low ESS of $p(\theta)$). The failure of \textit{real 2006} thus highlights the appeal of 1) the \textit{cat} or \textit{quant} parameterisations, both of which have trivial contributions to the prior density (ie. uniform priors), and 2) the smarter operators used by \textit{real} (Zhang and Drummond 2020). Due to its computational burden, \textit{real 2006} was not benchmarked for all of the datasets in \textbf{Table \ref{table:datasets}}.

\begin{figure}[!h]
%\includegraphics[width=\textwidth]{Figures/tournament.pdf}
\caption{\textbf{Rate parameterisation evaluation.} }
\label{fig:parameterisationResults}
\end{figure}

% 1-3) the \textit{cat}, \textit{real}, and \textit{quant} parameterisations using the operators described by Drummond et al. 2006 \cite{drummond2006relaxed}; 3) the \textit{quant} parameteristion
